Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-10-20 paper/1_Initial/manuscript.qmd
Update manuscript.qmd

<!-- UVA -->
After examining the distributions and correlations of all IAS items, we will test for "redundant" items (e.g., due to multicollinearity or local dependency) using Unique Variable Analysis [UVA, @christensen2023unique], a novel method derived from network psychometric designed to identify and merge items that share substantial variance (which can distort the structure estimation).
We will use a threshold of 0.30 that detects large to very large overlap.

Following the analysis of items, we will analyze the factor structure of the IAS using three different approaches, each with particular trade-offs and assumptions, to provide a multi-verse picture of likely solutions. Namely, we will apply traditional exploratory and confirmatory Factor Analysis (EFA/CFA), hierarchical clustering (HCA), and Exploratory Graph Analysis (EGA), to the whole sample, as well as to each dataset separately (details being available in the analysis document at https://github.com/RealityBending/InteroceptionIAS), and our decisions and conclusions will try to take into account both levels of analysis.

@@

```{r}
#| warning: false
#| label: fig-structure
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"
#| fig-cap: "TODO."

@@

<!-- Summary -->
Study 1 establishes that once probeleamtic items are removed the IAS is best conceptualized as a set of item clusters rather than a unidimensional scale. This structural insight provides the foundation for Study 2, where dispositional correlates are examined to further evaluate the IAS’s construct validity.
# Study 2

<!-- The goal of study 2 is two-fold, 1) provide a robust overview of the dispositional correlates of the IAS by leveraging the large number of datasets available, and 2) assess the practical benefits of using a refined version of the IAS (i.e., the 14-item version identified in study 1) compared to the original 21-item one general factor version. -->

## Study 2

Building on the structural findings from Study 1, the second study turns to the validity of the IAS by examining its dispositional correlates. Although the IAS has been widely used, evidence regarding its associations with personality traits, affective dispositions, and related psychological constructs remains scattered and inconsistent [@murphy2019; @brand2023; @lin2023; @todd2022]. Study 2 therefore had two primary goals. First, to provide a robust overview of the dispositional correlates of the IAS by leveraging the large number of available datasets. Second, to evaluate the practical benefits of using the refined 14-item version of the scale identified in Study 1, compared to the original 21-item unidimensional version, in order to assess whether structural improvements translate into clearer and more reliable external associations.

@@

We selected measures that appeared in multiple datasets from Study 1 and/or were relevant given the scope of the study, merging scores of regular and abridged versions for conciseness where applicable. The measures include constructs related to interoception, mood, personality, psychopathology, and beliefs and misbeliefs.

<!-- Interoception Related -->
#### Interoception Related
The Multidimensional Assessment of Interoceptive Awareness Version-2 [MAIA-2, @mehling2018multidimensional] is a 37-item questionnaire assessing eight dimensions of interoception: Noticing, Not-Distracting, Not-Worrying, Attention Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust. Responses are rated on a 6-point Likert scale (e.g., “I notice when I am uncomfortable in my body”).

>>>>

<!-- UVA -->
After examining the distributions and correlations of all IAS items, we will test for "redundant" items (e.g., due to multicollinearity or local dependency) using Unique Variable Analysis [UVA, @christensen2023unique], a novel method derived from network psychometric designed to identify and merge items that share substantial variance (which can distort the structure estimation).
We will use a threshold of 0.30 that detects large to very large overlap to remain conservative and only suggest scale modifications if strongly justified.

Following the analysis of items, we will analyze the factor structure of the IAS using three different approaches, each with particular trade-offs and assumptions, to provide a multi-verse picture of likely solutions. Namely, we will apply traditional exploratory and confirmatory Factor Analysis (EFA/CFA), hierarchical clustering (HCA), and Exploratory Graph Analysis (EGA), to the whole sample, as well as to each dataset separately (details being available in the analysis document at https://github.com/RealityBending/InteroceptionIAS), and our decisions and conclusions will try to take into account both levels of the analysis.

@@

```{r}
#| warning: false
#| label: fig-structure
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"
#| fig-cap: "Four structure analysis methods (HCA, EGA, EFA, CFA) were applied and converged on a consistent optimal solution of 14 items formed seven pairs: Hungry-Thirsty, Bruise-Blood sugar, Urinate-Defecate, Muscles-Pain, Breathing-Heart, Cough-Sneeze, Wind-Burp. While EFA suggested 3 factors, CFA confirmed the superiority of the 7-factor model over alternative structures."

@@

<!-- Summary -->
In summary, study 1 suggests that once problematic items are removed, the IAS is best conceptualized as a set of item clusters rather than a unidimensional scale. This structural insight provides the foundation for Study 2, where dispositional correlates are examined to further evaluate the IAS's construct validity.

# Study 2

Building on the structural findings from Study 1, the second study turns to the validity of the IAS by examining its dispositional correlates. Although the IAS has been widely used, evidence regarding its associations with personality traits, affective dispositions, and related psychological constructs remains scattered and inconsistent [@murphy2019; @brand2023; @lin2023; @todd2022]. Study 2 therefore had two primary goals. First, to provide a robust overview of the dispositional correlates of the IAS by leveraging the large number of available datasets. Second, to evaluate the practical benefits of using the refined 14-item version of the scale identified in Study 1, compared to the original 21-item unidimensional solution, in order to assess whether structural improvements translate into clearer and more reliable external associations.

@@

We selected measures that appeared in multiple datasets from Study 1 and/or were relevant given the scope of the study, i.e., constructs related to physiology, mood, personality, psychopathology, and beliefs and misbeliefs. We merged scores of regular and abridged versions for conciseness where applicable.

<!-- Interoception Related -->
#### Interoception Related

The Multidimensional Assessment of Interoceptive Awareness Version-2 [MAIA-2, @mehling2018multidimensional] is a 37-item questionnaire assessing eight dimensions of interoception: Noticing, Not-Distracting, Not-Worrying, Attention Regulation, Emotional Awareness, Self-Regulation, Body Listening, and Trust. Responses are rated on a 6-point Likert scale (e.g., “I notice when I am uncomfortable in my body”).
